\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-11-02 casestudies.tex
Venn diagram
istart on stlc case study

We

>>>>

To demonstrate the applicability of our design and implementation, we
conduct three case studies.

\noindentparagraph{Type safety of STLCs.}

The first case study is the mechanization of the type safety theorem of
STLC and those of its extensions,
which has been occurring in the examples in this paper.
The code base is ported from Software Foundations~\cite{sf-pl}.
%
The base STLC family consists of about 400 lines of code.
Lines of code in each of the four derived families
($\mathrm{Y}$, $\times$, $+$, and $\mu$ in the Venn diagram)
vary from 100 to 300, largely depending on how many constructors they
add to the inductive types.

\input{stlc-venn.tex}

Using individual families to organize the mechanization of individual
language features leads to a modular design that also facilitates code reuse.
%
Individually developed language features can be easily composed (as
mixins) to form new STLC variants.
%
Composing features can lead to \emph{feature interactions}~\cite{batory2011feature}:
features working correctly in isolation may require coordination when composed.
For example, composing \lsti{STLCIsorec} and \lsti{STLCProd}
(\cref{fig:stlc-isorec-prod}) creates a need to extend \lsti{tysubst} to
handle \lsti{ty_prod}, which the type-checker enforces.

The linguistic nature of our approach allows us to retain a programming
style similar to the original proofs while achieving extensibility.

### 2022-11-02 coq-example.tex
istart on stlc case study

\newpage\EDJ{Please delete this newpage command later}
We provide three examples here for illustration and evaluation of our Coq plugin.

\subsection{Type Safety for STLC}\EDJ{We need a fuller view of the two examples. Especially show "accessing family components outside that family".}
\label{sec:coqexample-stlc}

  \includegraphics[width=0.5\textwidth]{coqexmaple/Mixin-Venn-Diagram.pdf}

The first example is the example throughout our paper---type safety proof of STLC, largely adapted
from Software Foundation~\cite{pierce2014software}. We singled out each language extensions from
STLC and prove the type safety for them separately, following
the style of the examples in MTC~\cite{delaware2013,forsta2020}.
And we use the \textbf{mixin} feature---that is capable of mixing two linkage transformer---to mix the semantics and properties of language features---product
, boolean (in \cref{fig:STLC-example}), sum type and iso-recursive type---with vanilla STLC.

% Using this example,
% we hope one day we can say precisely, \textit{a programming language feature itself is a
% piece of data/inheritance judgement/inherited family}. 
Compared with MTC,
our example uses small-step operational semantics\YZ{Does MTC not define a small-step operational semantics for STLC?}\EDJreply{I thought they use big step with fuel. Though Coq a la carte use small-step in one example} by exploiting
extensible inductive types. Most of the proof and computation are directly adapted from the one in Software Foundation, resulting in a similarly accessible proof. However, we highlight some special expriences when using our plugin.

@@

We handle \textbf{inversion lemma} differently. Inversion lemmas are those small lemmas while involving global reasoning. One example is "not_step_tm_true" which trivially asserts "tm_true" is not further reducible. It requires global reasoning because once the user errorneously add a constructor for "step tm_true tm_true" then this lemma cannot hold. Thus one way is to prove it by "FInduction" on the "step".  However, this brings a lot of boilerplate code as everytime "step" is extended with some unrelated constructor (e.g. "st_if : step (tm_if tm_true b c) b"), "not_step_tm_true" needs to be extended correspondingly---we need to check "st_if" doesn't "step tm_true" requiring using \textbf{partial recursor and the computational axiom} to prove "(tm_if tm_true b c) <> tm_true" syntactically speaking using tactic "fdiscriminate".

Thus we provide a small-scale \textbf{escape hatch via "Overridable"}. Once "pins {step}", we will know the \textit{concrete} (vanilla Coq) inductive type of "step" when proving "not_step_tm_true" and thus we can use all Coq's induction facility including "inversion". Once the definition of "step" extends during inheritance, due to the semantic of "Overridable", we need to provide a new proof but our plugin can also try to use the old proof script automatically. This trick can apply to small inversion lemma where we believe our inductive definition is ``well-formed'' enough and thus an "inversion" tactic can easily handle it. Another way to look at it is the fact that, this kind of inversion lemma should be part of the definition in the first hand, like a constraint---and the proof that our inductive definition satisfy this constraint should be trivial and any future extension should respect this constraint.

@@

Our formulation for the base STLC family is around 400+ LOCs; each families implementing a single feature takes around 100~300 LOCs.

@@

\subsection{Abstract Interpreters for \texttt{Imp}}\YZ{Try to have two runnable interpreter -- choose another simple one like constant propagation. and remove LangMore}
\label{sec:coqexample-analysis}
The second example is adapted and modified from \citet{zm2017}.
Contrary to our first example, here we use a big-step interpreter with fuel
as the operational semantics of an imperative language with
side effects, and we specify the abstract interpretation and prove its
soundness. 
% Then we instantiate our abstract interpretation with two concrete computation---one for type inference and the other for constant propagation. 
Our analyzer will return the analysis result about the end of the program, e.g. our constant propagator will return what constant value each variable is at the end of the program. Thanks to the compilation to Coq module, we can directly run the resulting abstract interpreters.

@@

\begin{minted}[fontsize=\footnotesize]{Coq}
  Lemma analyze_faithful : forall prog st end_st, 
    interp prog st = end_st -> abs_interp prog (lift st) = (lift end_st)
\end{minted}

We construct four families, illustrated in \cref{fig:abstract-interpretation-example}.
The first family "Lang" define the big step operator (using fuel) of a
while language, with expression and stmt defined upon expression. Final "Result" can be error, "out-of-fuel" or a legitmate "State". It takes about 200 LOCs.

The second family "LangwAbs" define the abstract interpreter for "Lang" in "LangwAbs.AbsInterpreter".
The idea of abstract interpretation is to lift concrete interpretation, their value and their states into their abstract correspondence. Then we come up with a faithful (one example would be "analyze_faithful") and generic abstract intepreter. Then, the user can freely choose the instantiation and lifting of this generic framework and end up with their aiming program analyzer. For example, lifting 1 into abstract value "integer" can give us a type inferer; lifting 1 into an element of a very flat lattice can gives us a constant propagator. There are also special mathematical structure required in the abstract value (like lattices) for the abstract interpretation.

Thus, to implement this generic abstract interpreter, we need to postulate the abstract value and their mathematical property. Then we need to prove the abstract interpreter is \textbf{sound and terminating}  with the help of these postulations. Then for concrete analyzer, we only need to instantiate these postulation. The construction and proof of soundness uses "FInduction" and "FRecursion". "analyze_soundness" is saying if 
It takes about 300+ LOCs. 
% This example shows how
% to use family inheritance to extend a concrete interpreter with abstract
% interpreters. 
% This example also shows in this family polymoprhism framework we can reason about "Lang" with computation details being abstracted.

@@

The third family and the fourth family "LangTyping" and "LangCP" instantiate the postulates of the
"LangwAbs" and recover computational information of "analyze". The corresponding fields in "absStateData" is overridden by concrete implementations. The resulting abstract interpreter is expected
to act as a type-analyzer and constant-propagator respectively. It takes about 200+ LOCs. 
% These are two examples of instantiate the detail computation of abstract interpretation.

>>>>

\newpage
We provide three examples here for illustration and evaluation of our Coq plugin.

\subsection{Type Safety for STLC}
\label{sec:coqexample-stlc}

% \includegraphics[width=0.5\textwidth]{coqexmaple/Mixin-Venn-Diagram.pdf}

The first example is the example throughout our paper---type safety proof of STLC, largely adapted
from Software Foundation~\cite{pierce2014software}. We singled out each language extensions from
STLC and prove the type safety for them separately, following
the style of the examples in MTC~\cite{delaware2013,forsta2020}.
And we use the \textbf{mixin} feature---that is capable of mixing two linkage transformer---to mix the semantics and properties of language features---product
, boolean, sum type and iso-recursive type---with vanilla STLC.

% Using this example,
% we hope one day we can say precisely, \textit{a programming language feature itself is a
% piece of data/inheritance judgement/inherited family}. 
Compared with MTC,
our example uses small-step operational semantics\YZ{Does MTC not define a small-step operational semantics for STLC?}\EDJreply{I thought they use big step with fuel. Though Coq a la carte use small-step in one example} by exploiting
extensible inductive types. Most of the proof and computation are
directly adapted from the one in Software Foundation, resulting in a
similarly accessible proof. However, we highlight some special
experiences when using our plugin.

@@

We handle \textbf{inversion lemma} differently. Inversion lemmas are those small lemmas while involving global reasoning. One example is "not_step_tm_true" which trivially asserts "tm_true" is not further reducible. It requires global reasoning because once the user erroneously add a constructor for "step tm_true tm_true" then this lemma cannot hold. Thus, one way is to prove it by "FInduction" on the "step".  However, this brings a lot of boilerplate code as every time "step" is extended with some unrelated constructor (e.g. "st_if : step (tm_if tm_true b c) b"), "not_step_tm_true" needs to be extended correspondingly---we need to check "st_if" doesn't "step tm_true" requiring using \textbf{partial recursor and the computational axiom} to prove "(tm_if tm_true b c) <> tm_true" syntactically speaking using tactic "fdiscriminate".

Thus we provide a small-scale \textbf{escape hatch via "Overridable"}. Once "pins {step}", we will know the \textit{concrete} (vanilla Coq) inductive type of "step" when proving "not_step_tm_true" and thus we can use all Coq's induction facility including "inversion". Once the definition of "step" extends during inheritance, due to the semantic of "Overridable", we need to provide a new proof, but our plugin can also try to use the old proof script automatically. This trick can apply to small inversion lemma where we believe our inductive definition is ``well-formed'' enough and thus an "inversion" tactic can easily handle it. Another way to look at it is the fact that, this kind of inversion lemma should be part of the definition in the first hand, like a constraint---and the proof that our inductive definition satisfy this constraint should be trivial and any future extension should respect this constraint.

@@

Our formulation for the base STLC family is around 400+ LOCs; each family implementing a single feature takes around 100~300 LoCs.

@@

\subsection{Abstract Interpreters for \texttt{Imp}}\YZ{Try to have two runnable interpreter -- choose another simple one like constant propagation. and remove LangMore}
\label{sec:coqexample-analysis}
The second example is adapted and modified from \citet{zm2017}.
Contrary to our first example, here, we use a big-step interpreter with fuel
as the operational semantics of an imperative language with
side effects, and we specify the abstract interpretation and prove its
soundness, with some postulation on both computation and property. Then we instantiate our abstract interpretation with two concrete computation---one for type inference and the other for constant propagation. Our analyzer will return the analysis result about the end of the program, e.g. our constant propagator will return what constant value each variable is at the end of the program. Thanks to the compilation to Coq module, we can directly run the resulting
abstract interpreters.

@@

We construct four families, illustrated in \cref{fig:abstract-interpretation-example}.\YZ{Why is it useful to define these concrete and abstract interpreters as an inheritance chain?}\EDJreply{I don't think these examples are here to show the example themselves are useful. They are here to show family inheritance is useful, and can extend stuff to all sorts of things. I also add some details below on how each family can be extended and what each family means.}\YZreply{Why is it desired to define the abstract interpreter as an extension of the concrete interpreter family? What is lost by putting the two interpreters in the same family?}\EDJreply{Nothing is lost. This is just simulating people starting with bare (big-step) operational semantic. Because they may not working on abstract interpretation from the first place. Those people may extend "Lang" with other stuff like another type-safety proof or something.}
The first family "Lang" define the big step operator (using fuel) of a
while language, with expression and stmt defined upon expression. Final "Result" can be error, "out-of-fuel" or a legitmate "State". It takes about 200 LOCs. 
% We
% can
\YZ{In general, in a result section like this, you want to show what
you *have done* rather than what you *can do*.}\EDJreply{Ok. deleted. I initially want to emphasize ``I **can** use arbitrary implementation for this environment mapping interface'' but I don't know how to say that without using ``can''.}
% swap the implementation of "State" to have different memory/state
% management by extending family "Lang".
\YZ{What is 'different
memory/state management'?}\EDJreply{Actually I should say environment mapping/dictionary. I instantiate it with function(closure) as environment mapping in "LangRun". It is possible to use other mapping like Coq's primitive dictionary.}

The second family "LangwAbs" define the abstract interpreter for "Lang" in "LangwAbs.AbsInterpreter",
together with a bunch of postulates as a big sealed family "absStateData". Then we prove the correctness of the
abstract interpretation function "analyze". It takes about 300+ LOCs. This example shows how
to use family inheritance to extend a concrete interpreter with abstract
interpreters. 
% This example also shows in this family polymoprhism framework we can reason about "Lang" with computation details being abstracted.

@@

The third family and the fourth family "LangTyping" and "LangCP" instantiate some of the postulates of the
"LangwAbs" and recover computational information of "analyze". The corresponding fields in "absStateData" is overridden by a concrete definition. The resulting abstract interpreter is expected
to act as a type-analyzer and constant-propagator respectively. It takes about 200+ LOCs. 
% These are two examples of instantiate the detail computation of abstract interpretation.
